---
title: "607_project_2"
author: "Eddie Xu"
date: "2023-03-04"
output: html_document
---
## Project 2 - Tidying Up Data for Downstream Analysis
The goal of this assignment is to give you practice in preparing different datasets for downstream
analysis work

## Load Packages
```{r}
library(tidyverse)
library(maps)
library(mapdata)
```

## Load Datasets from different sources to seperate dataframes
```{r}
# From Nick Climaco's NAFTA country population
country_url <- ("https://raw.githubusercontent.com/Nick-Climaco/Rdataset/main/nafta_countries.csv")
country_file <- data.frame(read.csv(country_url, header = TRUE))

# From Genesis Middleton county GDP data
county_url <-("https://raw.githubusercontent.com/eddiexunyc/607_project_2/main/Resources/gdp_by_county.csv")
county_file <- data.frame(read.csv(county_url, header = TRUE))

# Interest Rate Data
interest_rate_url <- ("https://raw.githubusercontent.com/eddiexunyc/607_project_2/main/Resources/interest_rate.csv")
interest_file <- data.frame(read.csv(interest_rate_url, header = TRUE, sep = ";"))
```

## View all datasets
```{r}
# view the country dataframe
head(country_file)
```

```{r}
# view the county GDP dataframe
head(county_file)
```

```{r}
# view the interest rate dataframe
head(interest_file)
```

## Data Cleaning on the NAFTA Population Dataset
```{r}
# rename the column 
colnames(country_file) <- c("ID", "Country", "1994", "1995", "1996", "1997", "1998", "1999", "2000")

# pivot longer the year and population
country_tidy_df <- country_file %>%
  pivot_longer(c("1994", "1995", "1996", "1997", "1998", "1999", "2000"), names_to = "Year", values_to = "Population")

knitr::kable(country_tidy_df)
```

## Data Analysis on the NAFTA Population Dataset
```{r}
# change the class of Year to numeric
country_tidy_df$Year <- as.integer(country_tidy_df$Year)

# calculate the population difference and rate of change
growth_rate <- country_tidy_df %>%
  group_by(Country) %>%
  arrange(Year) %>%
  mutate(pop_difference = Population - lag(Population),
         rate_percent = round((pop_difference/lag(Population)) * 100, digits = 2))

# rename the column
colnames(growth_rate) <- c("ID", "Country", "Year", "Population", "Population Difference", "Population % Change")

knitr::kable(growth_rate)
```

```{r}
# graph a line chart of rate change in population
growth_bar <- ggplot(data = subset(growth_rate, !is.na(`Population % Change`)), aes(x = Year, y = `Population % Change`, fill = `Country`)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = `Population % Change`), position = position_dodge(0.9), color = "black", vjust = 1, hjust = 0.5)

growth_bar
```
## Conclusion in the NAFTA Population Dataset
Based on the graph, it shows that while Mexico have the highest growth rate among them all, the rate is getting smaller over time. Meanwhile, the US growth rate remains steady.

## Data Cleaning on the County GDP Dataset
```{r}
# remove first 2 rows
county_tidy_df <- county_file[-(1:2),]

# remove last 3 rows
county_tidy_df <- county_tidy_df[-(12453:12458),]

# rename the column
colnames(county_tidy_df) <- c("FIPS", "County Name", "State", "Line Code", "Industry Name", "GDP in 2012 (thousands of dollars)", "GDP in 2013 (thousands of dollars)", "GDP in 2014 (thousands of dollars)", "GDP in 2015 (thousands of dollars)")

head(county_tidy_df, 10)
```

## Data Analysis on the County GDP Dataset
```{r}
# change the GDP column from char type to numeric
county_tidy_df[6:9] <- lapply(county_tidy_df[6:9],as.numeric)

# calculate the sum and average of all GDPs per county
industry_df <- county_tidy_df %>%
  mutate(industry_total = rowSums(county_tidy_df[6:9], na.rm = TRUE),
         industry_average_gdp = round(rowMeans(county_tidy_df[6:9], na.rm =TRUE), digits = 2),
         High_GDP = apply(county_tidy_df[6:9], 1, max, na.rm = TRUE),
         Low_GDP = apply(county_tidy_df[6:9], 1, min, na.rm = TRUE))

# rename the column
colnames(industry_df)[10] <- "Industry Total"
colnames(industry_df)[11] <- "Industry Avg GDP"
colnames(industry_df)[12] <- "Highest GDP in the County"
colnames(industry_df)[13] <- "Lowest GDP in the County"

head(industry_df)
```

```{r}
# summarize the industry GDP per State
industry_plot_data <- subset(industry_df, `Industry Name` == "All Industries") %>%
  group_by(`State`) %>%
  summarise("Total GDP" = sum(`Industry Total`),
            "Highest GDP" = max(`Highest GDP in the County`),
            "Lowest GDP" = min(`Lowest GDP in the County`),
            )
# convert state abbreviations to name with a function [link]("https://rdrr.io/cran/usdata/src/R/abbr2state.R")

abbr2state <- function(abbr){
  ab    <- tolower(c("AL",
             "AK", "AZ", "KS", "UT", "CO", "CT",
             "DE", "FL", "GA", "HI", "ID", "IL",
             "IN", "IA", "AR", "KY", "LA", "ME",
             "MD", "MA", "MI", "MN", "MS", "MO",
             "MT", "NE", "NV", "NH", "NJ", "NM",
             "NY", "NC", "ND", "OH", "OK", "OR",
             "PA", "RI", "SC", "SD", "TN", "TX",
             "CA", "VT", "VA", "WA", "WV", "WI",
             "WY", "DC"))
  st    <- c("alabama",
             "alaska", "arizona", "kansas",
             "utah", "colorado", "connecticut",
             "delaware", "florida", "georgia",
             "hawaii", "idaho", "illinois",
             "indiana", "iowa", "arkansas",
             "kentucky", "louisiana", "maine",
             "maryland", "massachusetts", "michigan",
             "minnesota", "mississippi", "missouri",
             "montana", "nebraska", "nevada",
             "new hampshire", "new jersey", "new mexico",
             "new york", "north carolina", "north dakota",
             "ohio", "oklahoma", "oregon",
             "pennsylvania", "rhode island", "south carolina",
             "south dakota", "tennessee", "texas",
             "california", "vermont", "virginia",
             "washington", "west virginia", "wisconsin",
             "wyoming", "district of columbia")
  st[match(tolower(abbr), ab)]
}

industry_plot_data["Full Name"] <- abbr2state(industry_plot_data$State)

head(industry_plot_data, 10)
```

```{r}
# graph a map of the industry GDP
state <- map_data("state")
industry_map <- merge(state, industry_plot_data, by.x = "region", by.y = "Full Name")
industry_map <- arrange(industry_map, group, order)

ggplot(industry_map, aes(x = long, y = lat, group = group, fill = `Total GDP`)) +
  geom_polygon(colour = "black")
```

## Conclusion on the County GDP Dataset
In conclusion, it shows the total GDP in all industries in US and it appears that California have the highest GDP among all states.

## Data Cleaning on the Interest Rate Dataset
```{r}
# rename a column
colnames(interest_file) <- c("Date", "Federal Reserve System", "European Central Bank", "Swiss National Bank", "Bank of England", "Reserve Bank of Australia", "Bank of Japan", "Bank of Brazil")

head(interest_file)
```

## Data Analysis on the Interest Rate Dataset
```{r}
# filter the interest rate starting from the 2000
twentyfirst_century_df <- interest_file[interest_file$Date >= "2000-01-01" & interest_file$Date <= "2022-12-31", ]
twentyfirst_century_df$Date <- as.Date(twentyfirst_century_df$Date)

# pivot longer the central banks and interest rate
interest_rate_in_21stdata <- twentyfirst_century_df %>%
  pivot_longer(c( "Federal Reserve System", "European Central Bank", "Swiss National Bank", "Bank of England", "Reserve Bank of Australia", "Bank of Japan", "Bank of Brazil"), names_to = "Central Bank Name", values_to = "Interest Rate")

head(interest_rate_in_21stdata)

```

```{r}
ggplot(interest_rate_in_21stdata, aes(x = `Date`, y = `Interest Rate`, color = `Central Bank Name`)) +
  geom_point() +
  geom_hline(yintercept = 2, color = "red")
```

## Conclusion on the Interest Rate Dataset
In conclusion, the interest rate in 21st century shows the economic situation of every central bank's home country. For economy to be in good health, the interest rate usually need to be in in the range of 0 to 2%. Given that, Bank of Brazil have high interest rate to combat their economic uncertainty. Other central banks such as Bank of England and European Central Bank have been dropping their interest rates which show economic improvement.

